---
title: "Piecewise Regression"
author: "Bruna Wundervald"
output:
  rmarkdown::html_vignette:
    fig_width: 5
    fig_height: 3.5
    fig_cap: TRUE
    toc: yes
---
<style type="text/css">
#TOC {
  margin: 0 130px;
  width: 425px;
}
</style>

```{r setup, include = FALSE}
require(knitr)
opts_chunk$set(
    dpi = 100,
    fig.align = "center",
    comment = NA,
    message = FALSE,
    warning = FALSE,
    error = FALSE)
```

## Introduction

  - When analyzing the relationship between a response variable $Y$, 
  and some explanatory variables $\boldsymbol{X}$, it may happen that, 
  for different ranges of $\boldsymbol{X}$, the linear behaviour
  between the variables differs. In this situation, the use of more 
  than one linear model to describe the observed relationship might
  be interesting. In other words, it might be the case where the 
  application of *Piecewise Regression* is appropiate, because it 
  allows multiple linear models being fit fot the data, for 
  varying ranges of the explanatory variables. 
  
### How do we find the *breakpoint*?
  - Breakpoints are fundamental when using Piecewise Regression, 
  because they are points in the range of $X$ where the behaviour of
  $Y$ changes, then the name *"breakpoint"*. In some cases, the 
  breakpoint can be known, but normally it isn't. When this happens, 
  we must have a way to estimate its value.
  - Breakpoint estimation


### Different Breakpoints

```{r}
library(segmented)
library(shiny)
library(lattice)
library(latticeExtra)

da <- read.csv("df02.txt", 
               sep = "\t")

str(da)

print(xyplot(Y1 ~ x, data = da,
       type = c("p", "g", "smooth"), pch = 16,
             strip = strip.custom(bg = "white"),
       col = "tomato",
       xlab = "", ylab = ""),
     position = c(0,0, 0.5, 1), more = TRUE)

print(xyplot(Y2 ~ x, data = da,
             type = c("p", "g", "smooth"), pch = 16,
             strip = strip.custom(bg = "white"),
             col = "aquamarine",
             xlab = "", ylab = ""),
      position = c(0.5,0, 1, 1), more = TRUE)
```
_______________________________________________________________________

```{r, echo = FALSE, fig.width= 10, eval = FALSE}
#-------------------------------
bootstrapPage(div(style="display:inline-block",
                  sliderInput("bins", "Breakpoint 1", min = min(da$x),
                              max = max(da$x), value = 1,
                              step = 0.05)),
              div(style="display:inline-block",
                  sliderInput("bins2", "Breakpoint 2", min = min(da$x),
                              max = max(da$x), value = 1,
                              step = 0.05)))

output$chart <- renderPlot({
  da$v <- with(da,
               ifelse(x <= input$bins, 0, x - input$bins))
  da$p <- with(da,
               ifelse(x <= input$bins2, 0, x - input$bins2))
  m0 <- lm(Y1 ~ x + v, da)
  m1 <- lm(Y2 ~ x + p, da)
  
  da$pred <- predict(m0)
  da$pred2 <- predict(m1)
  aic1 <- AIC(m0)
  aic2 <- AIC(m1)
  #-----------
  par(mfrow = c(1, 2))
  plot(Y1 ~ x, data = da, pch = 16, main = paste0("AIC = ", 
                                                  round(aic1, 3)))
  da <- da[order(da$x),]
  with(subset(da, x < input$bins), lines(x, pred, col = "2", lwd = 2))
  with(subset(da, x >= input$bins), lines(x, pred, col = "2", lwd = 2))
  
  plot(Y2 ~ x, data = da, pch = 16, main = paste0("AIC = ", 
                                                  round(aic2, 3)))
  da <- da[order(da$x),]
  with(subset(da, x < input$bins2), lines(x, pred2, col = "2", 
                                           lwd = 2))
  with(subset(da, x >= input$bins2), lines(x, pred2, col = "2",
                                            lwd = 2))
})
plotOutput("chart")
```

<iframe src="https://brunaw.shinyapps.io/seg_shiny/" width=100% height =600 frameborder="0" scrolling="no" marginheight="0" marginwidth="0"> </iframe>


### Estimating the Breakpoint
  - For the estimation of the best breakpoint, we'll test every value
  between a certain interval and look to the *LogLikelihood* of each 
  adjusted model. The breakpoint that gives the bigger value for the
  LogLikelihood can be considered "the best one":
```{r}
#------------------
g <- numeric(length = 20)

ll <- for (i in 1:20) {
  m <- i/10
  da$v <- with(da, ifelse(x <= m, 0, x - m))
  m1 <- lm(Y1 ~ x + v, data = da)
  g[i] <- AIC(m1)
}
which.min(g) # 1.9
#-----------------
# Using the segmented package
s <-lm(Y1 ~ x, data = da)
o <- segmented(s, seg.Z=~x)

summary(o)
#----------------
# Second Dataset
#---------------
g <- numeric(length = 20)

ll <- for (i in 1:20) {
  m <- i/10
  da$v <- with(da, ifelse(x <= m, 0, x - m))
  m1 <- lm(Y2 ~ x + v, data = da)
  g[i] <- AIC(m1)
}
which.min(g) # 1.5
#-----------------
# Using the segmented package
s <-lm(Y2 ~ x, data = da)
o <- segmented(s, seg.Z=~x)
o$psi[2]
```


## Model Expression

  - When we have only one breakpoint, the model can be represented by:

$$ \begin{align}
y = \beta_{0} + \beta_{1} x , & \quad x \leq c \\
y = \beta_{0} + \beta_{2} x + \theta (\beta_{1}-\beta_{2}), & \quad x > c
\end{align} $$

## Types of Segments
  - We can also vary the behaviour of the line segments. Above, there is
  some examples of different situations:
  
### Different inclinations

$$ \begin{align}
y = \beta_{0} + \beta_{1} x , & \quad x \leq c \\
y = \beta_{0} + \beta_{2} x , & \quad x > c
\end{align} $$

```{r}
m <- lm(Y1 ~ x, data = da)
da$pred <- predict(m)

#----------------------
plot(Y1 ~ x, data = da)
da <- da[order(da$Y1),]
with(subset(da, x < 1), lines(x, pred, col = "2"))
da$sec <- m$coefficients[2] + 8 * da$x
with(subset(da, x >= 1), lines(x, sec, col = "2"))
#-----------------------
```

### Different intercepts

$$ \begin{align}
y = \beta_{0} + \beta_{1} x , & \quad x \leq c \\
y = \beta_{2} + \beta_{1} x , & \quad x > c
\end{align} $$

```{r}
plot(Y1 ~ x, data = da)
da <- da[order(da$x),]
with(subset(da, x < 1), lines(x, pred, col = "2"))
da$sec <- 1 + m$coefficients[2] * da$x
with(subset(da, x >= 1), lines(x, sec, col = "2"))
``` 

### Different intercepts and no inclination 

$$ \begin{align}
y = \beta_{0}, & \quad x \leq c \\
y = \beta_{1}, & \quad x > c
\end{align} $$

```{r}
plot(Y1 ~ x, data = da)
da <- da[order(da$x),]
da$c <- rep(15, nrow(da))
da$b <- rep(25, nrow(da))
with(subset(da, x < 1), lines(x, c, col = "2"))
with(subset(da, x >= 1), lines(x, b, col = "2"))
```


------------------------------------------------------------------------

## Model Comparison
  1) Linear Regression
  2) Transformed Response Variable
  3) Piecewise Regression
  4) Polinomial Regression
  5) Non-Linear Regression
  6) Generalized Linear Regression
  
```{r, fig.height=7, fig.width=7}
m1 <- lm(Y1 ~ x, data = da)
m2 <- lm(log(Y1) ~ x, data = da)

da$v <- with(da, ifelse(x <= 1.9, 0, x - 1.9))
m3 <- lm(Y1 ~ x + v, data = da)
m4 <- lm(Y1  ~ poly(x, 2), data = da)

m6 <- glm(Y1 ~ x, family = "Gamma"(link = "inverse"),
          data = da)

#------------------------
pred <- data.frame(m1 = predict(m1), m2 = exp(predict(m2)),
                   m3 = predict(m3), m4 = predict(m4),
                   m6 = predict(m6, type = "response"))
pred$x <- da$x
pred <- pred[order(pred$x),]


xyplot(Y1 ~ x,
       pch = 16,
       jitter.x = TRUE,
       type = c("p", "g"),
       xlab = expression(x[1]),
       ylab = list(label = "y",
                   rot = 0),
       scales = list(y = list(tick.number = 8)),
       da, col = 1,
       key = list(space = "bottom",
                  cex.title = 1.15,
                  text = list(c(
                    paste0("Linear Regression - aic = ", 
                           round(AIC(m1), 3)),
                    paste0("Transformed Response Variable - aic = ", 
                           round(AIC(m2), 3)),
                    paste0("Piecewise Regression - aic = ", 
                           round(AIC(m3), 3)),
                    paste0("Polinomial Regression - aic = ", 
                           round(AIC(m4), 3)),
                    paste0("Generalized Linear Regression - aic = ",
                           round(AIC(m6), 3)))),
                  lines =
                    list(col = c("yellow",
                                 "springgreen", "steelblue",
                                 "violetred", "tomato"),
                               lwd = 3))
       )+
  as.layer(xyplot(m1 + m2 + m3 + m4 + m6  ~ x ,
                  data = pred, type = "l",
                  auto.key = TRUE,
                  col = c("yellow", "springgreen", "steelblue",
                          "violetred", "tomato"),
                  lwd = 2))

#-------------------------------------------------------------
# Outlier Exclusion - for convergence of the Non-Linear Model
#------------------------------------------------------------

m1 <- lm(Y1 ~ x, data = da[-21,])
m2 <- lm(log(Y1) ~ x, data = da[-21,])

da$v <- with(da, ifelse(x <= 0.7, 0, x - 0.7))
m3 <- lm(Y1 ~ x + v, data = da[-21,])
m4 <- lm(Y1  ~ poly(x, 2), data = da[-21,])

# Incomplete Gamma Model
m5 <- nls(Y1 ~ (a*(x^b))*exp(-c*x), 
          start = c(a = 10, b = 0.01, c = -0.9),
          data = da[-21,])

m6 <- glm(Y1 ~ x, family = "Gamma"(link = "inverse"),
          data = da[-21,])

#------------------------
pred <- data.frame(m1 = predict(m1), m2 = exp(predict(m2)),
                   m3 = predict(m3), m4 = predict(m4),
                   m5 = predict(m5), 
                   m6 = predict(m6, type = "response"))
pred$x <- da[-21,]$x
pred <- pred[order(pred$x),]


xyplot(Y1 ~ x,
       pch = 16,
       jitter.x = TRUE,
       type = c("p", "g"),
       xlab = expression(x[1]),
       ylab = list(label = "y",
                   rot = 0),
       scales = list(y = list(tick.number = 8)),
       da, col = 1,
       key = list(space = "bottom",
                  cex.title = 1.15,
                  text = list(c(
                    paste0("Linear Regression - aic = ", 
                           round(AIC(m1), 3)),
                    paste0("Transformed Response Variable - aic = ", 
                           round(AIC(m2), 3)),
                    paste0("Piecewise Regression - aic = ", 
                           round(AIC(m3), 3)),
                    paste0("Polinomial Regression - aic = ", 
                           round(AIC(m4), 3)),
                    paste0("Non-Linear Regression - aic = ", 
                           round(AIC(m5), 3)),
                    paste0("Generalized Linear Regression - aic = ",
                           round(AIC(m6), 3)))),
                  lines =
                    list(col = c("yellow", "turquoise",
                                 "springgreen", "steelblue",
                                 "violetred", "tomato"),
                               lwd = 3))
       )+
  as.layer(xyplot(m1 + m2 + m3 + m4 + m5 + m6  ~ x ,
                  data = pred, type = "l",
                  auto.key = TRUE,
                  col = c("yellow", "turquoise", 
                          "springgreen", "steelblue",
                          "violetred", "tomato"),
                  lwd = 2))

```

**The AIC of the transformed response variable model is not comparable
to others as it is.**


## Using more than one breakpoint

  - When we have two breakpoints, for example, the model can be
  represented by:

$$ \begin{align}
y = \beta_{0} + \beta_{1} x_{1} + \beta_{2} x_{2} + \beta_{3} x_{3},
\\
\\
x_{2}=\left\{
  \begin{array}{ll}
    x_{1}-c_{1}, \quad c_{1} \leq x\\
    0, cc. \\
    \end{array}
\right.
\\
x_{3}=\left\{
    \begin{array}{ll}
    x_{1}-c_{2}, \quad c_{2} \leq x\\
    0, cc. \\
    \end{array}
\right.
\end{align} $$



```{r}
#----------------------------------------------------------------
da$x2 <- with(da, ifelse(x < 0.5, 0, x-0.5))
da$x3 <- with(da, ifelse(x < 0.5 | x >= 1.3, 0, x-1.3))
da$x4 <- with(da, ifelse(x >= 1.3, 0, x-1.3))

m1 <- lm(Y1 ~ x + x2 + x3, data = da) # With interval 
m2 <- lm(Y1 ~ x + x2 + x4, data = da) # Without interval 
logLik(m1); logLik(m2)


plot(Y1 ~ x, data = da, 
     main = paste0("Model with the interval - loglik = ", 
                   round(logLik(m1), 4)))
da$pred1 <- predict(m1)
with(subset(da, x < 0.5), lines(x, pred1, col = "2", lwd = 2))
with(subset(da,  0.5 > x | x <= 1.3), 
     lines(x, pred1, col = "2", lwd = 2))
with(subset(da, x >= 1.3), lines(x, pred1, col = "2", lwd = 2))

plot(Y1 ~ x, data = da, main = 
       paste0("Model without the interval - loglik = ", 
              round(logLik(m2), 3)))
da$pred2 <- predict(m2)
with(subset(da, x < 0.5), lines(x, pred2, col = "2", lwd = 2))
with(subset(da,  0.5 > x | x <= 1.3), 
     lines(x, pred2, col = "2", lwd = 2))
with(subset(da, x >= 1.3), lines(x, pred2, col = "2", lwd = 2))
```
 